---
title: "2026-03-15_p8106-hw02"
author: "Chris"
date: "2026-03-15"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
    css: style.css
---


```{r message=FALSE}
library(tidyverse)
set.seed(202603)
```



# Read and Split the Data

```{r}
df_college <- read_csv("data/College.csv") |>
  janitor::clean_names()

n <- nrow(df_college)
train_id <- sample(1:n, 0.8 * n)

df_train <- df_college[train_id, ]
df_test <- df_college[-train_id, ]
```


# Question a

## Fit the Models

```{r}
vec_df_grid <- 2:15

vec_x_grid <- seq(min(df_train$top10perc),
                  max(df_train$top10perc),
                  length.out = 200)
vec_curve_list <- vector("list", length(vec_df_grid))

for(i in seq_along(vec_df_grid)){
  df_one <- vec_df_grid[i]
  
  fit <- smooth.spline(x = df_train$top10perc,
                       y = df_train$outstate,
                       df = df_one)
  
  pred <- predict(fit, x = vec_x_grid)
  
  vec_curve_list[[i]] <- tibble(
    Top10perc = pred$x,
    Outstate_hat = pred$y,
    df = factor(df_one)
  )
}

df_curve_data <- bind_rows(vec_curve_list)
```


## Make the Plots

```{r}
plot1_qa <- ggplot() +
  geom_point(data = df_train,
             aes(x = top10perc, y = outstate),
             color = "grey60", alpha = 0.6) +
  geom_line(data = df_curve_data,
            aes(x = Top10perc, y = Outstate_hat, color = df),
            linewidth = 1) +
  labs(x = "Top10perc",
       y = "Outstate",
       color = "Degrees of Freedom",
       title = "Smoothing spline fits for different degrees of freedom") +
  theme_minimal()

plot1_qa

ggsave("plots/2026-03-15_plot1-question1.png", plot1_qa)
```


```{r}
plot2_qa <- ggplot() +
  geom_point(data = df_train,
             aes(x = top10perc, y = outstate),
             color = "grey60", alpha = 0.6) +
  geom_line(data = df_curve_data,
            aes(x = Top10perc, y = Outstate_hat, color = df),
            linewidth = 1) +
  facet_wrap(~ df, ncol = 4) +
  labs(x = "Top10perc",
       y = "Outstate",
       title = "Smoothing spline fits for different degrees of freedom") +
  theme_minimal()

plot2_qa
ggsave("plots/2026-03-15_plot2-question1.png", plot2_qa)
```


* As the degrees of freedom increase, the smoothing spline becomes more flexible.  
* For very small degrees of freedom (e.g df = 2 or 3), the fitted curve is nearly linear and captures only the overall increasing trend between `Top10perc` and `Outstate`.  
* As the degrees of freedom increase (df from f to 8), the curve begins to capture nonlinear patterns in the data, particularly slight curvature in the middle range of `Top10perc`.
* For larger degrees of freedom (\(df \ge 10\)), the fitted curves become more flexible and show additional local fluctuations, indicating that the model may start fitting noise.  
* However, the overall trend remains similar, suggesting that increasing the degrees of freedom beyond a certain point does not substantially change the fitted relationship.  


## Choose the Best DF

```{r}
fit_one <- smooth.spline(x = df_train$top10perc, y = df_train$outstate, cv = TRUE)
fit_one$df

pred_cv <- predict(fit_one, x = vec_x_grid)
df_cv_data <- tibble(
   Top10perc = pred_cv$x,
   Outstate_hat = pred_cv$y,
   df = fit_one$df
)

plot3_q1 <- ggplot() +
  geom_point(data = df_train, aes(x = top10perc, y = outstate),
             color = "red", alpha = 0.6) +
  geom_line(data = df_cv_data, aes(x = Top10perc, y = Outstate_hat), 
            color = "#6a5acd", linewidth = 1.2) +
  labs(x = "Top10perc",
       y = "Outstate",
       title = "Smoothing spline fits for the best degrees of freedom from cross validation") +
  theme_minimal()

plot3_q1
ggsave("plots/2026-03-15_plot3-question1_splines-for-cv.png", plot3_q1)
```



Cross-validation selected an effective degree of freedom of approximately `r fit_one$df`.  
































































